### 2011-07-29 docs/pubs/EIM-2011/application.tex
use pdf for figure 5 with larger font
incorporate comments from Reviewer #1

As discussed above, by leveraging the relationships defined and
inferred from the ontology we are able to increase recall beyond what
is possible for simple keyword-based searches
\cite{berkley09:_improv_data_discov_for_metad}. Broad queries return
direct matches as well as subclass matches. The queries can be quickly
refined when using the Web application by allowing rapid exploration
of the data repository without having to define complete observational
queries \emph{de novo}. The interface allows users to specify
individual classes of a measurement as well as pre-configured
measurement types (representing standard data-set attribute types) as
defined in OBOE compatible ontologies to enable a single concept to
proxy its constituent parts, namely the characteristics of particular
entities that can be measured with a set of protocols and
standards. This short-hand query generation can save users time in
specifying their queries, and highlights a compelling reason for using
OBOE extension ontologies. Measurement templates can also be leveraged when creating or
editing semantic annotations in the Morpho interface.

@@

\mypara{Data integration.} The materialization routine used for
semantic data queries laid the groundwork for enabling data
integration. In addition to inspecting the data for values within a
range and returning the data sets that contain a match, the data
integration feature of the Web application goes further by
constructing a synthetic data product (table) that represents the
complete results of the query in terms of both the attributes and the
values that are returned. Each original data set may have very
different syntactic structures (e.g. column number, naming, order) but
could share a subset of attributes that are semantically compatible as
defined in accompanying annotations. These compatible, in-common
attributes become the facets and metrics in a synthetic data set; their
values filtered accordingly. \figref{fig:integration} further
illustrates the data integration support provided in the current
implementation of the Web application. Consider two data packages,
denoted by A and B in \figref{fig:integration}.  Annotations (denoted
C and D) are used to determine semantically equivalent data attributes
contained in the data sets (denoted by E and F). The attributes {\tt
  plot} and {\tt site} are considered equivalent measurements of the
characteristic Location; attributes {\tt weight} and {\tt wt} both map
to the same characteristic Mass. The Semantic Mediation API parses
each annotation and then computes an equivalence mapping among
attributes based on their corresponding annotations. The Data Manager
Library is then used to load the data sets and then query each data
set to produce and merge the final, synthetic result data set.

@@

\begin{figure}
\centering
\includegraphics[scale=1.0]{images/integration.png}
\caption{Data integration query across multiple data packages (A, B).
  Annotations (C, D) are used to determine semantically equivalent
  data attributes contained in the data objects (E, F). Attributes
  {\tt plot} and {\tt site} are considered equivalent measurements of the
  characteristic Location; attributes {\tt weight} and {\tt wt} both map to
  the same characteristic Mass. The Semantic Mediation API utilizes
  the Data Manager Library to load and query the source data informed
  by semantic similarities between the structurally disparate data
  attributes.}
\label{fig:integration}
\end{figure}

>>>>

As discussed above, by leveraging the relationships defined and
inferred from the ontology we are able to increase recall beyond what
is possible for simple keyword-based searches
\cite{berkley09:_improv_data_discov_for_metad}. Broad queries return
direct matches as well as subclass matches. The queries can be quickly
refined when using the Web application by allowing rapid exploration
of the data repository without having to define complete observational
queries \emph{de novo}. The interface allows users to specify
individual classes of a measurement as well as pre-configured
measurement types (representing standard data set attribute types) as
defined in OBOE compatible ontologies to enable a single concept to
proxy its constituent parts, namely the characteristics of particular
entities that can be measured with a set of protocols and
standards. This short-hand query generation can save users time in
specifying their queries, and highlights a compelling reason for using
OBOE extension ontologies. Measurement templates can also be leveraged when creating or
editing semantic annotations in the Morpho interface.

@@

\mypara{Data integration.} The materialization routine used for
semantic data queries laid the groundwork for enabling data
integration. In addition to inspecting the data for values within a
range and returning the data sets that contain a match, the data
integration feature of the Semtools Web application goes further by
constructing a synthetic data product (table) that represents the
complete results of the query in terms of both the attributes and the
values that are returned. Each original data set may have very
different syntactic structures (e.g. column number, naming, order) but
could share a subset of attributes that are semantically compatible as
defined in accompanying annotations. These compatible, in-common
attributes become the facets and metrics in a synthetic data set; their
values filtered accordingly. \figref{fig:integration} further
illustrates the data integration support provided in the current
implementation of the Web application. Consider two data packages,
denoted by A and B in \figref{fig:integration}.  Annotations (denoted
C and D) are used to determine semantically equivalent data attributes
contained in the data sets (denoted by E and F). The attributes {\tt
  plot} and {\tt site} are considered equivalent measurements of the
characteristic Location; attributes {\tt weight} and {\tt wt} both map
to the same characteristic Mass. The Semantic Mediation API parses
each annotation and then computes an equivalence mapping among
attributes based on their corresponding annotations. The Data Manager
Library is then used to load the data sets and then query each data
set to produce and merge the final, synthetic result data set.

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{images/integration.pdf}
\caption{Data integration query across multiple data packages (A, B).
  Annotations (C, D) are used to determine semantically equivalent
  data attributes contained in the data objects (E, F). Attributes
  {\tt plot} and {\tt site} are considered equivalent measurements of the
  characteristic Location; attributes {\tt weight} and {\tt wt} both map to
  the same characteristic Mass. The Semantic Mediation API utilizes
  the Data Manager Library to load and query the source data informed
  by semantic similarities between the structurally disparate data
  attributes.}
\label{fig:integration}
\end{figure}

### 2011-07-29 docs/pubs/EIM-2011/framework.tex
use pdf (vector) graphic for figure 2 and let latex do the scaling. Font is increased to 12 (mostly)
incorporate comments from Reviewer #1

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.49\textwidth]{images/oboe}
  \caption{The main classes and properties of the extensible
    observation ontology (OBOE). While shown using UML, the model is
    defined using OWL-DL.}
  \label{fig:oboe}
\end{figure}

@@

\begin{figure}[!b]
\centering
\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
\caption{Partial OBOE semantic annotation for Kelp sampling
  data. Shaded nodes represent ontological concepts; rectangular nodes
  are data table attributes mapped to OBOE measurement
  characteristics.}
\label{fig:kelp-mass-model}
\end{figure}

@@

Semantic annotations can be used to facilitate discovery and
integration of heterogeneous data sets. For instance, combining
semantic annotations with OBOE, it is possible to discover data sets
based on searches expressed over types of observations and
measurements of interest. As simple examples, users can pose queries
such as ``\emph{find all data sets containing observations of Kelp
  entities}'' and ``\emph{find all data sets containing Mass
  measurements of Kelp}''.\footnote{Note that queries also can include
  contextual constraints, e.g., Mass measurements of Kelp within a
  specific spatial or temporal context.} Both of these queries would
return the example data set in \figref{fig:kelp-mass-model} since the
attribute {\tt WET} is linked to a WetMass measurement for observations
of Macrocystis (where WetMass is defined as a special kind, or
\emph{subclass} of Mass, and Macrocystis is defined as a subclass of
Kelp). Using semantic annotations in this way can help to increase
both query recall and precision over standard keyword-based approaches
\cite{berkley09:_improv_data_discov_for_metad}. In particular, by
defining terms as subclasses of other terms (e.g., Macrocystis as a
subclass of Kelp), term expansion can be used to increase the number
(recall) of data sets returned (where subclasses of query terms are
also searched). The precision of the result can be improved since
queries may specify the desired connections between terms (e.g.,
measurements of Mass for Kelp observations) as opposed to returning
all data sets that simply mention the terms but without any explicit
connections (i.e., where Mass was measured, but not for Kelp samples).
Annotations also help facilitate integration by allowing tools to
align data-set attributes based on their declared measurement and
observation types.

@@

\mypara{The Semantic Mediation API.} The Semantic Mediation API
includes basic ontology management features, annotation manipulation
capabilities, and simple concept navigation and visualization
components. The API is intended to be a centralized toolkit for use in
multiple application contexts (on either client or server
deployments).  The Semantic Mediation API uses both the OWL API
\cite{owlapi} for ontology management services including RDF/OWL
parsing, serialization, and simple class and property exploration as
well as the Pellet description-logic reasoner \cite{pellet} for
classification and exposing inferred axioms in source ontologies. The
inference services exposed through the Semantic Mediation API are used
in both discovery and integration features described below.
% is particularly powerful when utilizing OBOE Measurement 'templates'
% that express exactly what (Entity and Characteristic) can be
% observed and how (Standard and Protocol).
In our current Morpho and Metacat extensions, semantic annotations are
managed and stored automatically in an underlying, local relational
database. While it is also possible to use in-memory approaches for
storing and querying annotations, we found the overhead to be
prohibitive when large numbers of data sets are managed.

@@

\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
plugin for Morpho provides a front-end to the Semantic Mediation API
and allows data owners and curators to define annotations for existing data package
descriptions. The editor provides a simple ``fill-in-the-blank'' style
form-based interface with a searchable hierarchical concept selection
widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
integrates with a standard Morpho installation and provides semantic
query capabilities for locating data packages, marking up data
packages with semantic annotations, and saving annotations locally or
to a shared repository where they can be discovered and explored by
other users. The annotation editor in Morpho allows a user to view the
data set being annotated as they fill in (by selecting an appropriate
ontology term) the characteristic, measurement standard, protocol, and
associated entity for each data-set attribute. Users can also specify
whether an observation spans multiple columns, and can provide context
relationships between attributes (i.e., observations). The editor
provides a number of additional features including the ability to
view the entire annotation (similar to \figref{fig:kelp-mass-model})
and to specify additional mapping constraints for observations and
measurements.

>>>>

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.49\textwidth]{images/oboe}
  \caption{The main classes and properties of the extensible
    observation ontology (OBOE). While shown using Unified Modeling Language (UML), the model is
    defined using OWL-DL.}
  \label{fig:oboe}
\end{figure}

@@

\begin{figure}[!b]
\centering
\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.pdf}
\caption{Partial OBOE semantic annotation for Kelp sampling
  data. Shaded nodes represent ontological concepts; rectangular nodes
  are data table attributes mapped to OBOE measurement
  characteristics.}
\label{fig:kelp-mass-model}
\end{figure}

@@

Semantic annotations can be used to facilitate discovery and
integration of heterogeneous data sets. For instance, combining
semantic annotations with OBOE, it is possible to discover data sets
based on searches expressed over types of observations and
measurements of interest. As simple examples, users can pose queries
such as ``\emph{find all data sets containing observations of Kelp
  entities}'' and ``\emph{find all data sets containing Mass
  measurements of Kelp}''.\footnote{Note that queries also can include
  contextual constraints, e.g., Mass measurements of Kelp within a
  specific spatial or temporal context.} Both of these queries would
return the example data set in \figref{fig:kelp-mass-model} since the
attribute {\tt WET} is linked to a WetMass measurement for observations
of Macrocystis (where WetMass is defined as a special kind, or
\emph{subclass} of Mass, and Macrocystis is defined as a subclass of
Kelp). Using semantic annotations in this way can help to increase
both query recall and precision over standard keyword-based approaches
\cite{berkley09:_improv_data_discov_for_metad}. In particular, by
defining terms as subclasses of other terms (e.g., Macrocystis as a
subclass of Kelp), term expansion can be used to increase the number
(recall) of data sets returned (where subclasses of query terms are
also searched). The precision of the result can be improved since
queries may specify the desired connections between terms (e.g.,
measurements of Mass for Kelp observations) as opposed to returning
all data sets that simply mention the terms but without any explicit
connections (i.e., where Mass was measured, but not for Kelp samples).
Annotations also help facilitate integration by allowing tools to
align data set attributes based on their declared measurement and
observation types.

@@

\mypara{The Semantic Mediation API.} The Semantic Mediation API
includes basic ontology management features, annotation manipulation
capabilities, and simple concept navigation and visualization
components. The API is intended to be a centralized toolkit for use in
multiple application contexts (on either client or server
deployments).  The Semantic Mediation API uses both the OWL API
\cite{owlapi} for ontology management services including ontology 
parsing, serialization, and simple class and property exploration as
well as the Pellet description-logic reasoner \cite{pellet} for
classification and exposing inferred axioms in source ontologies. The
inference services exposed through the Semantic Mediation API are used
in both discovery and integration features described below.
% is particularly powerful when utilizing OBOE Measurement 'templates'
% that express exactly what (Entity and Characteristic) can be
% observed and how (Standard and Protocol).
In our current Morpho and Metacat extensions, semantic annotations are
managed and stored automatically in an underlying, local relational
database. While it is also possible to use in-memory approaches for
storing and querying annotations, we found the overhead to be
prohibitive when large numbers of data sets are managed.

@@

\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
plugin for Morpho provides a front-end to the Semantic Mediation API
and allows data owners and curators to define annotations for existing EML data 
descriptions. The editor provides a simple ``fill-in-the-blank'' style
form-based interface with a searchable hierarchical concept selection
widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
integrates with a standard Morpho installation and provides semantic
query capabilities for locating data packages, marking up data sets within a 
package using semantic annotations, and saving annotations locally or
to a shared repository where they can be discovered and explored by
other users. The annotation editor in Morpho allows a user to view the
data set being annotated as they fill in (by selecting an appropriate
ontology term) the characteristic, measurement standard, protocol, and
associated entity for each data set attribute. Users can also specify
whether an observation spans multiple columns, and can provide context
relationships between attributes (i.e., observations). The editor
provides a number of additional features including the ability to
view the entire annotation (similar to \figref{fig:kelp-mass-model})
and to specify additional mapping constraints for observations and
measurements.

### 2011-07-29 docs/pubs/EIM-2011/intro.tex
incorporate comments from Reviewer #1

In this paper we describe extensions to the Ecological Metadata
Language (EML) \cite{Fegraus07} and supporting tools for enabling
improved discovery and integration of ecological data sets. Our work
is based on the Extensible Observations Ontology (OBOE)
\cite{bowers08,madin07:_ontol_for_descr_and_synth}, which represents a
generic observational model implemented in OWL-DL for describing
domain-specific observation and measurement types. Our approach adds
additional metadata in the form of semantic annotations that link
attributes within data sets to OBOE terms for describing the implicit
observation and measurement types found within data sets. Semantic
annotations are executable in the sense that they can be used to
convert a data set into a collection of observation and measurement
instances, providing a more uniform representation for expressing
queries and performing integration.  To support the creation of
annotations, we have extended the Morpho metadata editor
\cite{metacat02:_manag_heter_ecolog_data_using_morph} with a
high-level user interface as well as the Metacat data catalog
\cite{berkley01:_metac} for storing and querying annotations through a
new Semantic Mediation API. This API can also be used to perform basic
data-level integration tasks using our prior work on the EML Data
Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin}.

>>>>

In this paper we describe extensions to the Ecological Metadata
Language (EML) \cite{Fegraus07} and supporting tools for enabling
improved discovery and integration of ecological data sets. Our work
is based on the Extensible Observations Ontology (OBOE)
\cite{bowers08,madin07:_ontol_for_descr_and_synth}, which represents a
generic observational model implemented in OWL-DL \cite{owldl} for describing
domain-specific observation and measurement types. Our approach adds
additional metadata in the form of semantic annotations that link
attributes within data sets to OBOE terms for describing the implicit
observation and measurement types found within data sets. Semantic
annotations are executable in the sense that they can be used to
convert a data set into a collection of observation and measurement
instances, providing a more uniform representation for expressing
queries and performing integration.  To support the creation of
annotations, we have extended the Morpho metadata editor
\cite{metacat02:_manag_heter_ecolog_data_using_morph} with a
high-level user interface as well as the Metacat data catalog
\cite{berkley01:_metac} for storing and querying annotations through a
new Semantic Mediation API. This API can also be used to perform basic
data-level integration tasks using our prior work on the EML Data
Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin}.
